\chapter{Glossary}
\label{app:glossary}

\begin{description}[style=nextline,leftmargin=!,labelwidth=2.2cm]
  \item[APL] A Programming Language; array-oriented notation used at Layer~1.
  \item[ASP] Answer Set Programming; non-monotonic reasoning at Layer~7.
  \item[Bifrost] The WORM chain recording every step across the constellation.
  \item[C{--}] A portable, typed assembly language; the Layer~5 target.
  \item[Certificate] A SHA-256 of a transformation rule plus a structural
        Merkle root, attesting a rewrite.
  \item[Coherence] The constellation-level predicate (entropy + intercoil)
        gating advancement.
  \item[Ed25519] An Edwards-curve digital signature scheme; the Plasma Gate.
  \item[FFI] Foreign-function interface; the C{--}$\leftrightarrow$Lean bridge.
  \item[FOL] First-order logic; the Layer~6 resolution prover.
  \item[IR] Intermediate representation; the canonical S-expression (Layer~2).
  \item[Liquid Haskell] Refinement-typing for Haskell (Layer~3).
  \item[OP\_RETURN] A Bitcoin output type carrying arbitrary data;
        used for external timestamping.
  \item[Plasma Gate] The Ed25519-enforced entry control on every artifact.
  \item[P/NP swarm] Multi-agent witness search with P-time verification
        (Layer~8).
  \item[Receipt] A signed, hash-linked record of an artifact (Layer~9).
  \item[Refinement type] A base type constrained by a decidable predicate.
  \item[Safety boundary] A point at which one representation is transformed
        into the next under an explicit, checkable contract.
  \item[Sorry] A Lean placeholder admitting an open proof goal; tracked by
        \texttt{sorrydb}.
  \item[VSCP] Verified Symbolic Compute Pipeline; the ten-layer architecture.
  \item[WORM] Write-once-read-many; an append-only, tamper-evident ledger.
\end{description}
