\chapter{Bibliography}
\label{app:bibliography}

\begin{thebibliography}{99}

\bibitem{coq} The Coq Proof Assistant. \url{https://coq.inria.fr}.

\bibitem{isabelle} Naraschewski, W., and Nipkow, T. (1998).
\textit{Isabelle/HOL: A Proof Assistant for Higher-Order Logic}. Springer.

\bibitem{lean4} de Moura, L., and Ullrich, D. (2021).
\textit{The Lean 4 Theorem Prover and Programming Language}. ITP.

\bibitem{mathlib} The mathlib Community. (2020).
\textit{The Lean Mathematical Library}. CPP.

\bibitem{liquidhaskell} Vazou, N., et al. (2014).
\textit{LiquidHaskell: Haskell as a Theorem Prover}. Haskell Symposium.

\bibitem{compcert} Leroy, X. (2009).
\textit{Formal Verification of a Realistic Compiler}. CACM.

\bibitem{cmm} Ramsey, N., and Peyton Jones, S. (2000).
\textit{A Single Intermediate Language That Supports Multiple Implementations
of Exceptions}. POPL. (C{--}.)

\bibitem{apl} Iverson, K. E. (1962).
\textit{A Programming Language}. Wiley.

\bibitem{ct} Laurie, B., and Langley, A. (2013).
\textit{Certificate Transparency}. RFC 6962.

\bibitem{bitcoin} Nakamoto, S. (2008).
\textit{Bitcoin: A Peer-to-Peer Electronic Cash System}.

\bibitem{sigstore} Sigstore. \url{https://sigstore.dev}. (Transparency-log
software supply-chain signing.)

\bibitem{ed25519} Bernstein, D. J., et al. (2011).
\textit{High-speed high-security signatures}. CHES.

\bibitem{leanffi} The Lean FFI documentation. \url{https://lean-lang.org}.

\bibitem{clingo} Gebser, M., et al. (2014).
\textit{Clingo: A Grounder and Solver for Answer Set Programs}.

\end{thebibliography}
