\chapter{The Coherence and Governance Model}
\label{ch:coherence}

Before the pipeline is described, its operating context must be. \textsc{mathlib5}
does not run in isolation; it participates in a governed constellation of
repositories. This chapter explains the coherence model that decides whether
any component is permitted to advance.

\section{The Resonance Field}
\label{sec:coh-field}

The constellation publishes a ``Resonance Field'' snapshot (the
\texttt{OMEGA-FIELD} block in the root \texttt{README.md}). Two quantities
govern coherence:
\begin{itemize}
  \item \textbf{Entropy $E$} --- a Shannon-style disorder metric over the
        repository set. The field requires $E < 0.21$; the observed value is
        $E = 0.1183$, comfortably under threshold.
  \item \textbf{Intercoil connectivity} --- the number of linked memory-graph
        and Bifrost connections. Six memory-graph and seven Bifrost links are
        reported.
\end{itemize}
Coherence is a Prolog-style predicate:
\begin{verbatim}
coherent(system) :-
    entropy(E), E < 0.21,
    intercoil(_, memory_graph),
    intercoil(_, bifrost_engine).
\end{verbatim}

\section{The Governance Kernel}
\label{sec:coh-kernel}

A compact \textsc{apl} incantation summarizes the trust model:
\begin{verbatim}
\ensuremath{Omega} \ensuremath{leftarrow} \ensuremath{Box}\ensuremath{wedge}\ensuremath{circ}\ensuremath{wedge}\ensuremath{Diamond}\ensuremath{wedge}\ensuremath{triangle}\ensuremath{wedge}\ensuremath{hexagon}
\end{verbatim}
read as Source ($\bigodot$) $\ensuremath{rightarrow}$ Truth ($\boxdot$) $\ensuremath{rightarrow}$
Resonance ($\bigcirc$) $\ensuremath{rightarrow}$ Knowledge ($\ensuremath{Diamond}$) $\ensuremath{rightarrow}$
Creation ($\ensuremath{triangle}$) $\ensuremath{rightarrow}$ Impact ($\ensuremath{hexagon}$) $\ensuremath{rightarrow}$
Unified ($\ensuremath{Omega}$). The governing rule is explicit:

\begin{quote}
\itshape No component advances unless every previous component remains
coherent.
\end{quote}

This is not metaphor. It is a deployment gate: a build that raises entropy
above threshold, or that severs an intercoil link, is rejected. The pipeline's
own safety boundaries (Chapter~\ref{ch:vscp}) are a localized instance of the
same principle.

\section{The WORM Seal}
\label{sec:coh-seal}

The field carries an ``\ensuremath{Omega} WORM Seal'' --- a SHA-256 root such as
\texttt{859f93e1\ldots168790} --- recomputed on every update. Because the seal
is itself hash-linked, the entire published state is tamper-evident. Each
\textsc{mathlib5} artifact inherits this property through its own
\texttt{metadata.json} (see \cref{ch:worm}).

\section{The Plasma Gate}
\label{sec:coh-plasma}

Every component's security audit records a \texttt{plasma\_gate} of
\texttt{Ed25519\_Enforced}. The Plasma Gate is the cryptographic entry
control: no artifact is admitted to the ledger unless it carries a valid
Ed25519 signature from an authorized key. Combined with AES-256-GCM
encryption at rest and SHA-256-chain tamper evidence, the gate realizes the
``immutable system of safety boundaries'' named in this monograph's title.

\section{Implications for the Pipeline}
\label{sec:coh-impl}

Three consequences flow downstream into the VSCP:
\begin{enumerate}
  \item \textbf{Lineage is mandatory.} Because the Bifrost WORM chain records
        every step, each pipeline layer can cite its parent receipts.
  \item \textbf{Signing is non-negotiable.} Ed25519 enforcement means a layer
        cannot emit an artifact without attributing it.
  \item \textbf{Coherence is observable.} The entropy metric gives operators a
        single number for ``is the system still safe to advance.''
\end{enumerate}
With the governance context established, we turn to the pipeline itself.
