\chapter{Layer 1: Symbolic APL Frontend}
\label{ch:apl}

\section{Purpose and Contract}
\label{sec:apl-purpose}

Layer~1 is the human interface. Its inbound contract is ``a string the user
believes is an \textsc{apl} expression''; its outbound contract is ``a typed
abstract syntax tree in the expression language of \cref{sec:pre-apl}.'' The
layer must fail if the string is not grammatical.

\section{The Expression Language}
\label{sec:apl-expr}

The reference frontend (\texttt{layers/apl/src/Parser.hs}) defines a small but
expressive language:
\begin{lstlisting}[language=Haskell,caption={Expression datatype (Parser.hs).},label=lst:apl-expr]
data Expr
  = IntVal Int
  | FloatVal Double
  | Var String
  | Lambda [String] Expr
  | App Expr [Expr]
  | Prim String [Expr]
  deriving (Show, Eq)
\end{lstlisting}
This is a \emph{fragment} of \textsc{apl}: it captures integer and float
literals, variables, lambda abstraction (the \texttt{\{ \ensuremath{dots} \}} notation),
application, and primitive combinators. The full density of \textsc{apl} is
intentionally out of scope; the fragment is chosen so that every downstream
layer has a well-defined input.

\section{The Parser}
\label{sec:apl-parser}

Parsing uses Megaparsec over \texttt{Text}. The top-level combinator tries a
lambda, then an application, then a term:
\begin{lstlisting}[language=Haskell,caption={Top-level parser (Parser.hs).},label=lst:apl-parse]
parseAPL :: Text -> Either (ParseErrorBundle Text Void) Expr
parseAPL = parse (space *> pExpr <* eof) ""

pExpr :: Parser Expr
pExpr = pLambda <|> pApp <|> pTerm
\end{lstlisting}
The lambda parser treats the brace body as an expression bound to the
conventional \textsc{apl} argument \texttt{\ensuremath{omega}}:
\begin{lstlisting}[language=Haskell,caption={Lambda parsing (Parser.hs).},label=lst:apl-lambda]
pLambda :: Parser Expr
pLambda = do
  _ <- char '{'
  space
  e <- pExpr
  space
  _ <- char '}'
  return $ Lambda ["omega"] e
\end{lstlisting}
Application is left-recursive in spirit but implemented by gathering a head
term and a tail of terms:
\begin{lstlisting}[language=Haskell,caption={Application and term parsing (Parser.hs).},label=lst:apl-app]
pApp :: Parser Expr
pApp = do
  t  <- pTerm
  ts <- many pTerm
  if null ts then return t else return (App t ts)

pTerm :: Parser Expr
pTerm = choice
  [ IntVal . read <$> some digitChar
  , Var    <$> some letterChar
  , between (char '(') (char ')') pExpr
  ]
\end{lstlisting}

\section{Worked Example}
\label{sec:apl-example}

The repository's \texttt{examples/playground/simple\_sum.apl} reads:
\begin{lstlisting}[language=,caption={simple\_sum.apl.},label=lst:apl-ex]
# Sum of squares: Sum_{k=1}^n k^2
SumSquares <- { (+/ (iotaomega) * 2) }
# Expected closed form: n(n+1)(2n+1) / 6
\end{lstlisting}
After parsing, \texttt{SumSquares} is a \texttt{Lambda ["\ensuremath{omega}"] (App (Var "+/")
[App (App (Var "*") [App (Var "\ensuremath{iota}") [Var "\ensuremath{omega}"], IntVal 2])])} (modulo the exact
application nesting). This tree is the artifact handed to Layer~2.

\section{Maturity and Limits}
\label{sec:apl-limits}

The current frontend parses a tractable fragment. It does not yet implement
\emph{tacit} (operator-only) \textsc{apl}, trains of operators, or the full
set of primitive glyphs. These are explicit non-goals of the present
implementation; the boundary contract is honored for the supported fragment,
and input outside it is rejected rather than silently mis-parsed.

\section{Further APL Examples}
\label{sec:apl-further}

Beyond $\Sigma\,\omega$, representative accepted expressions include:
\begin{itemize}
  \item weighted sum $w+. \times\, v$ (dot product via outer product then
        reduction),
  \item matrix row contraction $(\omega \circ.\times M)[i;]$,
  \item cumulative prefix $+\backslash v$ (scan), handled by a dedicated
        IR node with its own closed-form certificate when $v$ is
        arithmetic.
\end{itemize}
Each is translated by the same Layer~1 $\to$ 2 contract, so the safety
argument carries over unchanged.
