\chapter{The Constellation and the Agent Substrate}
\label{ch:constellation}

\textsc{mathlib5} is one node in a much larger graph of repositories. This
chapter situates it and describes the runtime substrate that executes its
swarm layer.

\section{Scale}
\label{sec:con-scale}

The SNAPKITTYWEST constellation reports 93 repositories (the operator states
over 108 across orgs, some private). Of these, 82 were active within 30 days
and 33 publish GitHub Pages sites. The Resonance Field (\cref{ch:coherence})
tracks entropy across the set and declares the whole coherent at
$E = 0.1183$.

\section{Mathlib5's Siblings}
\label{sec:con-siblings}

Several sibling repositories are directly relevant to this monograph:
\begin{itemize}
  \item \texttt{math-skills} --- six sealed skills (enumeration, isomorphism,
        symmetry, Hadamard, probabilistic, circuits) with master seals; the
        production instance of Layer~8's PRISM skills.
  \item \texttt{mathrosetta}, \texttt{resonance-math}, \texttt{math-engine}
        --- complementary mathematics repositories.
  \item \texttt{collatz-verifier}, \texttt{axiom-proof}, \texttt{qec-discovery}
        --- verification and discovery efforts.
  \item \texttt{agentos} --- the P/NP swarm substrate that schedules the
        agents referenced in Layer~8.
\end{itemize}

\section{The Agent Substrate (AgentOS)}
\label{sec:con-agentos}

\texttt{agentos/} is a runtime for solver agents. Its design matches the
swarm protocol of \cref{sec:pnp-swarm}:
\begin{itemize}
  \item A \textbf{problem registry} lists open problems with P-time
        verifiers.
  \item Agents \textbf{claim} a problem (append-only claim ledger).
  \item Agents \textbf{solve} (NP-hard witness search) and \textbf{submit} to
        a solution pool.
  \item A verifier promotes the first verified solution; the
        \textbf{universe sum} (a monotonic convergence metric) advances.
\end{itemize}
This is the execution engine behind \texttt{layers/pnp-attack/}. The
``universe sum $\ensuremath{rightarrow} \ensuremath{infty}$'' goal is a poetic statement of the same
completeness metric introduced for proofs in \cref{ch:completeness}.

\section{The Governance Kernel, Again}
\label{sec:con-gov}

The constellation's governance kernel
\begin{quote}
\centering
$\ensuremath{Omega} \ensuremath{leftarrow} \ensuremath{Box} \ensuremath{wedge} \ensuremath{circ} \ensuremath{wedge} \ensuremath{Diamond} \ensuremath{wedge} \ensuremath{triangle} \ensuremath{wedge} \ensuremath{hexagon}$
\end{quote}
is not merely decorative. It is the policy that the ASP layer
(\cref{ch:policy}) compiles into executable constraints and that the
Resonance Field enforces at the repository level. \textsc{mathlib5}
participates in this governance by emitting receipts that the Bifrost WORM
chain records; the chain link \texttt{Bifrost\_WORM\_Chain\_20260710\_01}
appears in every component's \texttt{metadata.json}.

\section{Implications}
\label{sec:con-implications}

The practical implication is that no single repository is the system; the
system is the coherent, receipt-linked constellation. A verification in
\textsc{mathlib5} can cite receipts from \texttt{math-skills}; a swarm agent
in \texttt{agentos} can claim a problem defined in \textsc{mathlib5}. The
safety boundaries described in this monograph are local instances of a
global coherence discipline.

\section{The Constellation as a Documentation Graph}
\label{sec:const-docs}

Beyond runtime, Layer~9's metadata feeds a documentation graph: each
artifact links to its proofs, its receipts, and its downstream consumers.
A reader browsing the repository thus sees not prose but a navigable
constellation of certified facts, echoing the ``galaxy'' metaphor of the
AGENTS.md. The graph is regenerated on every verified build, so it can
never describe an artifact that was not actually proven.
