\chapter{Case Study: Sum of Squares, End to End}
\label{ch:case}

We trace a single query through all ten layers to make the architecture
concrete. The query is the sum of squares, $\sum_{k=1}^{n} k^2$.

\section{Layer 1 -- APL}
\label{sec:case-l1}

The user writes, in \texttt{examples/playground/simple\_sum.apl}:
\begin{lstlisting}[language=,caption={Input.},label=lst:case-in]
SumSquares <- { (+/ (iotaomega) * 2) }
\end{lstlisting}
The Megaparsec parser (\cref{ch:apl}) yields an \texttt{Expr} tree: a lambda
bound to \texttt{\ensuremath{omega}} whose body is an application of \texttt{+/} to
\texttt{(\ensuremath{iota}\ensuremath{omega})*2}.

\section{Layer 2 -- IR}
\label{sec:case-l2}

The tree is converted to canonical S-expression (\cref{ch:ir}):
\texttt{(+/ (* (\ensuremath{iota} \ensuremath{omega}) 2))} in the uniform \texttt{SApp} form. Canonicalization
ensures any equivalent spelling maps here.

\section{Layer 3 -- Refinement}
\label{sec:case-l3}

The Liquid bridge (\cref{ch:liquid}) renders the expression to Lean and emits
\texttt{theorem proof\_obligation : <expr> := by sorry}. The refinement
predicate \texttt{wellFormed} (specified, not yet enforced) would here confirm
no dangling variable beyond \texttt{\ensuremath{omega}}.

\section{Layer 4 -- Lean}
\label{sec:case-l4}

Lean receives the obligation. In a fully formalized build it would prove the
closed form; today it records a \texttt{sorry} in \texttt{sorrydb} and the
pattern registry supplies the closed form $\frac{n(n+1)(2n+1)}{6}$.

\section{Layer 5 -- C{\textendash}{\textendash}}
\label{sec:case-l5}

The kernel computes the closed form via \texttt{rewrite\_with\_cert} (or, for
the benchmark, \texttt{bridge\_rewrite} at pattern id 1), returning
$n(n+1)(2n+1)/6$ with a certificate slot filled.

\section{Layer 6 -- Logic/Static}
\label{sec:case-l6}

The FOL checker confirms the associated first-order statement; CodeQL confirms
the implementing code has no obvious defect class. Both pass.

\section{Layer 7 -- Policy}
\label{sec:case-l7}

ASP consults the seal: the artifact was produced by an authorized agent with a
valid receipt, so \texttt{may\_publish} holds in a stable model.

\section{Layer 8 -- Optimization/Swarm}
\label{sec:case-l8}

The closed form is cached as a PRISM skill (canonical JSON + SHA-256d + WORM
seal), so the next identical query reuses it without recomputation. No swarm
search is needed for this solved case.

\section{Layer 9 -- WORM}
\label{sec:case-l9}

Every artifact above --- the parsed AST, the IR, the Lean obligation, the
kernel result, the FOL report, the policy decision --- is receipted with a
SHA-256, hash-linked into the Bifrost chain, and Ed25519-signed.

\section{Layer 10 -- Completeness}
\label{sec:case-l10}

The \texttt{sorryhunter} attempts to close the remaining Lean gap by
cross-checking the kernel's closed form against Lean's independent
computation; agreement promotes the obligation from \texttt{sorry} toward
closed, raising the completeness score $C$.

\section{Takeaway}
\label{sec:case-take}

A one-line \textsc{apl} expression becomes, by the time it is answered, a
forest of receipts, each attributable and each checkable. That is the entire
thesis of this monograph in miniature.
