\chapter{Conclusion}
\label{ch:conclusion}

\section{Summary}
\label{sec:con-summary}

We presented \textsc{mathlib5}, a Verified Symbolic Compute Pipeline organized
as ten explicit safety boundaries. Each boundary has an inbound contract, a
transformation, and an outbound contract; each failure is localized,
reported, and sealed. The pipeline moves an \textsc{apl} expression through a
canonical IR, Liquid Haskell refinement, Lean~4 proof obligations, a verified
C{--} kernel, logic/static analysis, ASP policy reasoning, a P/NP optimization
swarm, WORM provenance receipts, and automated proof-completeness checking.

\section{What We Established}
\label{sec:con-established}

\begin{enumerate}
  \item A coherent layered model in which confidence is increased by making
        every transformation inspectable.
  \item A concrete, public, multi-language reference implementation whose
        sources are reproduced in the appendices.
  \item A verified numeric kernel in C{--} whose primitives each name their
        safety boundary and certificate.
  \item A foreign-function bridge that cross-validates the kernel against
        Lean~4, so two independent paths must agree.
  \item A provenance layer that hash-links, Ed25519-signs, and Bitcoin-anchors
        every artifact --- including this manuscript.
\end{enumerate}

\section{The Central Claim}
\label{sec:con-claim}

The claim of this work is modest but firm: \emph{confidence in mathematical
software is increased when transformations between symbolic notation and
executable artifacts remain inspectable, provable, and cryptographically
attributable.} The ten-layer pipeline is one realization of that claim; the
constellation of receipt-linked repositories is its ecosystem.

\section{Outlook}
\label{sec:con-outlook}

The active frontier is closing the gap between architecture and operation:
enforcing the Liquid predicates, populating \texttt{Core.lean} with closed
theorems, wiring the end-to-end orchestration, and hardening the kernel's
certificate checks. As \texttt{sorryhunter} closes goals, the completeness
score $C$ rises; as the swarm finds harder witnesses, the universe sum
advances. The system is designed so that every step of that progress is
recorded, signed, and tamper-evident.

\section{Final Word}
\label{sec:con-final}

The machine is the extension of the mind. \textsc{mathlib5} is an attempt to
make that extension \emph{accountable}: not faster math, but math you can
trace, prove, and trust --- sealed so that, should it ever be altered, the
alteration is visible to anyone who cares to look. The fortress holds. For
now, and by design, indefinitely.

\vspace{0.3in}
\begin{center}
\itshape
\ensuremath{odot} Source \ensuremath{cdot} \ensuremath{Box} Truth \ensuremath{cdot} \ensuremath{circ} Resonance \ensuremath{cdot} \ensuremath{Diamond} Knowledge \ensuremath{cdot} \ensuremath{triangle} Creation \ensuremath{cdot} \ensuremath{hexagon} Impact \ensuremath{cdot} \ensuremath{Omega} Unified \\
--- WORM SEALED
\end{center}

\section{Open Questions}
\label{sec:concl-open}

Several questions remain active: extending the APL fragment to
higher-rank arrays, proving closed forms for non-polynomial summands, and
tightening the SMT bounds in the Liquid layer. Each is tracked as a P/NP
problem (\cref{ch:pnp}) and as a \texttt{sorrydb} entry (\cref{ch:lean-meta}),
so progress is publicly observable on the Bifrost chain.
