\chapter{Comparison with Established Systems}
\label{ch:compare}

\section{Scope of the Comparison}
\label{sec:compare-scope}

To situate MATHLIB5, we compare it along four axes against systems that
share part of its mission: verified compilation (CompCert, CertiCoq),
proof assistants with large libraries (Lean \texttt{mathlib}, Isabelle,
Coq), and dependently-typed numeric kernels (Herbie, FPTaylor).

\begin{table}[ht]
\centering
\caption{Comparison of MATHLIB5 against established verification and
  numeric systems.}\label{tab:compare}
\small
\begin{tabular}{p{3.2cm} p{2.4cm} p{2.4cm} p{2.4cm} p{2.4cm}}
\toprule
\textbf{Dimension} & \textbf{CompCert} & \textbf{Lean mathlib} &
\textbf{CertiCoq} & \textbf{MATHLIB5} \\
\midrule
Target language & C & Lean & Gallina/Coq & APL (user) $\to$ C$--$ \\
Proof carrier & Coq & Lean & Coq & Lean4 + Liquid Haskell \\
Numeric guarantees & none & library-level & none & closed-form + bounds \\
Cross-language bridge & no & limited & Coq$\to$OCaml & yes (C/Lean) \\
On-chain provenance & no & no & no & yes (Bifrost) \\
Build-gate enforcement & no & no & no & yes (halts) \\
\bottomrule
\end{tabular}
\end{table}

\section{What MATHLIB5 Adds}
\label{sec:compare-add}

\begin{itemize}
  \item \textbf{An end-to-end pipeline}, not a single verified compiler.
        CompCert verifies C$\to$asm; CertiCoq verifies Gallina$\to$C;
        MATHLIB5 covers APL$\to$IR$\to$Liquid$\to$Lean$\to$C$--$, each
        boundary independently certifiable.
  \item \textbf{Closed-form + numeric correctness} as a first-class goal,
        whereas CompCert and CertiCoq are silent on floating-point
        semantics beyond IEEE-754 preservation.
  \item \textbf{Public provenance}. Only MATHLIB5 ties receipts to a
        Bitcoin anchor; the others rely on the integrity of their own
        repositories.
\end{itemize}

\section{What MATHLIB5 Does Not Claim}
\label{sec:compare-not}

We do not claim to supersede \texttt{mathlib} as a general mathematical
library, nor CompCert as a production C compiler. MATHLIB5 is a
\emph{domain-specific} verified compute substrate for summation and
matrix algebra, with a deliberately narrow but fully certified surface.

\section{Related Work in Verified Numerics}
\label{sec:compare-num}

Herbie and FPTaylor attack floating-point accuracy via rewriting and
interval/SMT analysis. MATHLIB5 complements them: where they optimize an
existing expression's accuracy, MATHLIB5 certifies that a closed-form
expression is the \emph{correct} one and that the C$--$ translation
preserves its value semantics. A future integration could feed Herbie's
rewrites through the same boundary chain.
